# File Input and Output
This lesson will cover some standard functions for reading and writing data in R.

## Objectives
- getting and setting working directory
- save and load R objects to/from disk
- read and write tabular data
- read data from a url

## Basic Idea
As a data scientist, you will constantly be reading from and writing to files.
Generally you are given some dataset that you need to analyze and report on.
This means that you need to load the data into R, run some code, and finally save some outputs.
This is all done with files. 

## File Formats
When people talk about binary files, vs text files what they really mean is - is it human readable?
A text file should have text data that a human can read with a text editor.
A binary file has binary data that a human can't really read, but the appropriate software can.

## Filesystems and Paths
At a high level, files are information stored on a computer.
Each file has a name, and a unique file path.
A filepath is the location of the file in the storage device.
A file name as two parts: the name and the file extension.

The file extension (everything after the .) is meant to indicate to the user (you) and the operating system what the file contains.
This hint means you often times don't need to open the file to know what kind of data it has.
However, the extension is not enforced by anything, its just a useful suggestion.

Paths can be relative or absolute. 
An absolute path is the full path through the filesystem to reach a file.
A relative path is the path from some starting point to the file you want to reach.
You can consider a relative path as something that needs to be combined with another path to reach a file.

##  get and set working directory 
Before we begin, we will get and set our working directory in R.
You can think of the working directory as the part that gets combined with the relative path.

`getwd()` will return the absolute file path of the working directory
```{r eval=FALSE}
getwd()
```


Call setwd to set the working directory to a path you specify as an input argument.
```{r eval=FALSE}
setwd("~/Documents/file_io/") # notice the argument
getwd()
```

A really useful function in R is list.files(), which lists all the files at a given path.
Listing the files should confirm for us that we are in the right place.
```{r eval=FALSE}
list.files() 
```


## saving and loading R data

### rds
A simple way to save an R object directly to a file, such that it can be loaded into another R session is with the `saveRDS` function. 
`saveRDS` will write a single object to a specified file path.
By default, it will save the object as a binary representation.
This can be very useful for large objects, as the binary format will be significantly more space efficient.
```{r eval=FALSE}
y = c(0,1,2,3,4)
saveRDS(y, file="myvectors.rds")
```

Confirm that it worked.
```{r eval=FALSE, message=FALSE, warning=FALSE}
list.files()
```

The counterpart to `saveRDS` is `readRDS`.
With `readRDS`, you can load in an rds file, which by definition contains a single R object, and assign it to a variable in your session.
```{r eval=FALSE}
x = readRDS("myvector.rds")
```

This will work for any R object.
For example. 
```{r eval=FALSE}
saveRDS(mtcars, file="mtcars.rds")
my_cars = readRDS("mtcars.rds")
```

Saving and loading using `readRDS` is really powerful to save data.
However, it does have a pretty significant drawback - its useless outside of R.
For someone to explore the data, they would need to load R.

## Reading and Writing Text data
In addition to `saveRDS` and `readRDS`,
R has functions for working with text files.

These are commonly used for getting external data into R. 
And for exporting your data so that it can be used by other people.

### tabular data
Generally the data you work with in R will be tabular.
Dataframes are an example of tabular data.

### read and write table
To write tabular data from a text file use the write.table function.
Before running it, lets look at the documentation and understand the key arguments.

```{r eval=FALSE}
?write.table 
```
The important arguments are x, file, and sep.

**x** is the dataframe you are saving.
**file** is the name of the file you want to create and write to.
**sep** is the field separator, also called delimiter.

Notice that if file is left blank, then R will just print the results to the console, instead of into a file.
Lets use this to explore what the **sep** argument does.

```{r eval=FALSE, message=FALSE, warning=FALSE}
small = head(my_cars)
write.table(small)
write.table(small, sep=" ")
write.table(small, sep=".")
write.table(small, sep=",")
```

Lets write our data to a file.
```{r eval=FALSE}
write.table(my_cars, file="cars.txt") 
list.files() # confirm it worked
```

Now lets read the data back in. 
```{r eval=FALSE}
from_cars.txt = read.table("cars.txt") 
```

Always inspect your data to make sure everything worked
```{r eval=FALSE}
colnames(from_cars.txt)
dim(from_cars.txt)
head(from_cars.txt)
```
### CSV format

A CSV (comma separated values) file is a text file that uses ',' as the field separator.
This is probably the most commonly used format for plain text tabular data.

To write a csv in R use the `write.csv` function.
This is equivalent to `write.table(from_cars.txt, file="cars.csv", sep=",")`
```{r eval=FALSE}
write.csv(from_cars.txt, file="cars.csv") 
from_cars.csv = read.csv("cars.csv")
```

Again, double check that everything worked.
```{r eval=FALSE}
head(from_cars.csv)
colnames(from_cars.csv)
dim(from_cars.csv)
```

What went wrong here?
In this case it was ambiguous if the first column was rownames or actual values.

Lets fix it
```{r eval=FALSE}
temp = from_cars.csv[, 2:12]
rownames(temp) = from_cars.csv[,1]
fixed = temp
```

With these sorts of problems, you can generally fix them by using the appropriate arguments to the function calls of the read and write functions.
Notice the argument in this function call. 
It specifies that the rownames can be read in from the first column of the tabular data in the file.
```{r eval=FALSE}
from_cars.csv = read.csv("cars.csv", row.names =1) 
```

### Non tabular data

There are functions in R for reading and writing text data that doesn't represent tabular data.
A common one is `writeLines` and `readLines`.
```{r eval=FALSE}
texts = c("line one", "line two")
writeLines(texts, "raw.txt")
texts2 = readLines("raw.txt")
```

## URLS as files
Files can be transferred over the internet. 
**URLs** are a type of filepath, that denotes a filepath, and the computer that file is stored on.
Many functions in R that involve reading and writing from files, can be given a url as the filepath argument.
In that case, the file will be transferred over the internet, onto your computer, and then read into R.

Here is an example of reading in a file from a url, using the `readLines` function.
```{r eval=FALSE}
url = "https://datalab.ucdavis.edu"
t = readLines(url)
```
