# Working with files in R (and packages)

This lesson will focus on working with data files in R. 
It will reinforce understanding of the command line, as well as RStudio.
And, it will demonstrate finding and loading packages in R.

## Objectives

After this lecture you should be able to:

- Identify file extensions
- Identify some common data science file formats
- Read csv files into R
- Save data to disk as RDS
- Download packages from CRAN
- Read excel files into R

## Setup

To follow along, download this zip file from the following url:
[https://datalab.ucdavis.edu/adventures-in-datascience/best_in_show.zip](https://datalab.ucdavis.edu/adventures-in-datascience/best_in_show.zip)

Navigate to where you want to save your work:
```
cd ~/Documents/
```

Next, make a directory:
```
mkdir files_in_r
```

```
cd files_in_r
```

Copy the downloaded zip file into that directory:

```
cp ~/Downloads/best_in_show.zip .
```

Unzip the file:
```
unzip best_in_show.zip
```

Navigate to the newly created directory
```
cd best_in_show
```

## Exploring Files

When working with files, its important to gather lots of information, and
constantly test assumptions that you may have.

This process is a key part of programming and of working in the command line.

Lets start by seeing what we have, which we do with the `ls` command:
```
ls
```

Remember that `ls` can be modified with flags, for example, to see *all* the 
files including hidden ones, use the `-a` flag:
```
ls -a
```

You can see more information about the files with the `-l` flag:
```
ls -l
```

Modifiers can be combined for `ls`:
```
ls -la
```

You can use `du -h` to see the disk usage (file size) of a given file:
```
du -h dogs.csv
```

`-h` refers to human readable as by default `du` displays the size in block
units. Being aware of the size of a file early on can help debug issues with
running out of disk space, as well as issues down the line in the analysis 
process. For example, reading in too many too large files into R can create
issues for you by overloading your system's RAM (your computer's working memory). 

You can view the disk usage for all the files in the directory with the 
wildcard symbol `*`:
```
du -h *
```

### File Extensions

Most of the time, you can guess the format of a file by looking at its
extension, the characters (usually three) after the last dot . in the filename.
For example, the extension .jpg or .jpeg indicates a JPEG image file. Some
operating systems hide extensions by default, but you can find instructions to
change this setting online by searching for “show file extensions” and your
operating system’s name. The extension is just part of the file’s name, so it
should be taken as a hint about the file’s format rather than a guarantee.

### Text Files

A text file is one that contains human-readable lines of text. You can check
this by opening the file with a text editor such as Microsoft Notepad or macOS
TextEdit. Many file formats use text in order to make the format easier to work
with.

On the command line, you can get information about the type of a file
by using the `file` command followed by the name of a file:
```
file dogs.csv
```

Note that `file` uses a series of tests (learn more by reading `man file`),
to determine the file type and may not always perfectly report the type of the 
file.

The output of `file` is the filename followed by a colon and then a description
of the file type. 

In this case, the output tells us that dogs.csv is a CSV text file.

A **comma-separated values** (CSV) file records tabular data
using one line per row, with commas separating columns. 

From the command line we can read text files with `vim`:
```
vim dogs.csv
```

To see the type of all the files in the directory you can use the wildcard `*` 
operator:
```
file *
```

### Binary Files

A binary file is one that’s not human-readable. You can’t just read off the
data if you open a binary file in a text editor, but they have a number of
other advantages. Compared to text files, binary files are often faster to read
and take up less storage space (bytes).

For demonstrations sake, see what happens when you try to use vim to 'read'
a binary data file:
```
vim dogs.rds
```

Notice that the editor displays data but it isn't human readable, it looks
like a bunch of random symbols with potentially the occasional recognizable 
word.

### Common Data File Types

| Name                        | Extension  | Tabular?   | Text?
| :-------------------------- | :--------  |  :-------- | :----
| Comma-separated Values      | `.csv`     |  Yes       | Yes
| Tab-separated Values        | `.tsv`     |  Yes       | Yes
| Fixed-width File            | `.fwf`     |  Yes       | Yes
| Microsoft Excel             | `.xlsx`    |  Yes       | No
| Microsoft Excel 1993-2007   | `.xls`     |  Yes       | No
| [Apache Arrow][arrow]       | `.feather` |  Yes       | No
| R Data                      | `.rds`     |  Sometimes | No
| R Data                      | `.rda`     |  Sometimes | No
| Plaintext                   | `.txt`     |  Sometimes | Yes
| Extensible Markup Language  | `.xml`     |  No        | Yes
| JavaScript Object Notation  | `.json`    |  No        | Yes

[arrow]: https://arrow.apache.org/


## Reading and Writing Files in R

R has many functions for working with file systems, reading and writing
files. 

### The Working Directory

The _working directory_ is the starting point R uses for relative paths. 
Think of the working directory as the directory R is currently "at" or watching.

The function `getwd` returns the absolute path for the current working
directory, as a string. It doesn't require any arguments:

```{r, eval=FALSE}
getwd()
```
On your computer, the output from `getwd` will likely be different. This is a
very useful function for getting your bearings when you write relative paths.
If you write a relative path and it doesn't work as expected, the first thing
to do is check the working directory.

The related `setwd` function changes the working directory. It takes one
argument: a path to the new working directory. Here's an example:

```{r, eval=FALSE}
setwd("..")
# Now check the working directory.
getwd()
```

Generally, you should avoid using calls to `setwd` in your R scripts and R
Markdown files. Calling `setwd` makes your code more difficult to understand,
and can always be avoided by using appropriate relative paths. If you call
`setwd` with an absolute path, it also makes your code less portable to other
computers. It's fine to use `setwd` interactively (in the R console), but avoid
making your saved code dependent on it.

When working in RStudio, you can set the working directory _at the start of your
session_ in `session` -> `Set Working Directory` -> `To Source File Location`

Another function that's useful for dealing with the working directory and file
system is `list.files`. The `list.files` function returns the names of all of
the files and directories inside of a directory. It accepts a path to a
directory as an argument, or assumes the working directory if you don't pass a
path. For instance:

```{r, eval=FALSE}
# List files and directories in ~/.
list.files("~/")
# List files and directories in the working directory.
list.files()
```

If you call `list.files` with an invalid path or
an empty directory, the output is `character(0)`:

```{r, eval=FALSE}
list.files("/this/path/is/fake/")
```

Later on, we'll learn about what `character(0)` means more generally.

### Reading Our CSV File

Let's go ahead and read the dogs.csv file we extracted from the zip file
at the start.

R provides a very easy built-in function for reading CSV files, and a variety
of other formats for text files containing tabular data.

To read a csv file into R, use `read.csv`:
```{r, eval=FALSE}
dogs = read.csv('dogs.csv')
```

#### inspect our data

Note that we used read.csv assuming that the file extension was appropriate,
we can feel confident about this because we looked into the file earlier.

We can also feel confident that it will fit into memory as we saw it was 
relatively small.

Whenever you import data into R, it is crucial to check that things went 
as expected.

To check things went according to our expections, look at the output
of the read.csv function which we saved into `dogs`.

Lets see what the output is. We can check what the object is with the `class`
function.
```{r, eval=FALSE}
class(dogs)
```

We can see that the read.csv function returned a dataframe. This makes sense
because dataframes represent tabular data, and csv files contain tabular data.

We can get more information with the `str` function. `str` concisely gives 
information about the content of an R object:
```{r, eval=FALSE}
str(dogs)
```

Let's check the dimensions of our dataset:
```{r, eval=FALSE}
dim(dogs)
```

Recall we can access the number of rows with:
```{r, eval=FALSE}
nrow(dogs)
```

And the number of columns:
```{r, eval=FALSE}
ncol(dogs)
```

To display the first rows from the dataset, use `head`:
```{r, eval=FALSE}
head(dogs)
```

And to display the last rows from the dataset, use `tail`:
```{r, eval=FALSE}
tail(dogs)
```

### read functions for tabular data

`read.csv` is actually shortcut for `read.table` a general purpose
function for reading tabular data from plain-text files into R.

`read.csv` acts just like `read.table` but sets the `sep = ","` and `header=TRUE`

Table of R's read functions (https://rstudio-education.github.io/hopr/dataio.html)

| Function      | Defaults                            | Use
| :--------     | :--------                           | :---
|`read.table`   |sep = " ", header = FALSE            | General-purpose read function
|`read.csv`     |sep = ",", header = TRUE             | Comma-separated-variable (CSV) files
|`read.delim`   |sep = "\t", header = TRUE            | Tab-delimited files
|`read.csv2`    |sep = ";", header = TRUE, dec = ","  | CSV files with European decimal format
|`read.delim2`  |sep = "\t", header = TRUE, dec = "," | Tab-delimited files with European decimal format


### RDS

You can save any R object, such as a data frame, as an RDS file.
RDS files are a great option for storing data that is intended to be loaded
into R. Data saved as RDS can be quickly and accurately loaded out of and
back into R without losing any information. 

This isn't always the case when saving data in plain text formats such as CSV.
Any R-related metadata associated with the object you are saving will be
maintainted in the RDS format. This is useful in the case of data frames if
your data contains factors, or dates, or other specific class attributes that
won't be represented in a csv. You would need to reproduce the process for
parsing the data into R.

Additionally, RDS files often times take significantly less disk space to save,
as they are a compressed format. RDS files in general are faster to read.

However, its important to keep in mind that RDS files are meant to be used
only in R. If you save data as an RDS, you are assuming that however is using
that data will have access to and an understanding of R. 

As a result, its common to use the RDS format for saving intermediary data
in a project. While when exporting results to a collaborator, or the internet
you would most likely want to use a commonly used plain-text format such as
CSV.

Use `saveRDS` to save our data as an rds file with the `rds` file extension.
```{r, eval=FALSE}
saveRDS(dogs, "./outputs/dogs.rds")
```

Its easy to load RDS files in R with `readRDS`:
```{r, eval=FALSE}
dogs = readRDS("./outputs/dogs.rds")
```

### Excel Files in R

Excel is very popular in the data analysis world. Millions of people use 
Excel to input, clean, analyze, and store data.


## Packages

Lot's of the most useful parts of R do not come preloaded when you install R.
Packages bundle together code, documentation and data. Its easy to share,
and easy to include in your own code. Users have contributed thousands
of R packages which can be found online.

You can think of a package as one or more functions that are related to a 
specific task, that you can include in your code.

Packages need to be installed on your system and then loaded into 
your R session.


### CRAN

**Comprehensive R Archive Network** (CRAN) is the main website that makes
R packages accessible. 

#### readxl

**readxl** is a package written to provide functions for working with xlsx
files in R.

### Using Packages 

To use an R package, it first needs to be installed on your system, 
and then loaded into the R session.

#### Installing packages

You can install packages from CRAN onto your system using `install.packages`. 
It will search for the package on CRAN, and download the code onto your computer
in a place that R can access. 

To install the `readxl` package, we pass the name to `install.packages`:
```{r, eval=FALSE}
install.packages("readxl")
```

#### Loading packages

Even if the package is on your system, it is not automatically loaded into R.

Every time you restart R you will need to reload each package that your script
uses. Do so with `library` at the top of your script for each package that 
you will use.

This signals to yourself and anyone else that uses your script which packages
are required to run the code, and will stop the execution of the script
if any of the packages are not found.

To load in the `readxl` package we installed in the previous step use `library`:
```{r, eval=FALSE}
library("readxl")
```

This will load in all the functions, data, and documentation from the `readxl`
library, so we can now access them in our R session.

To see all the packages installed you can run `library` without any arguments:
```{r, eval=FALSE}
library()
```

This displays all the installed libraries as well the path R is searching
to find them.

###  Load Excel Data 

With Readxl we can list all the sheets in an excel spreadsheet:
```{r, eval=FALSE}
sheets = excel_sheets("./data/dogs.xlsx")
```

And can load the data with `read_excel`
```{r, eval=FALSE}
data = read_xlsx("./data/dogs.xlsx")
```
