# Indexing
This lesson will cover some standard functions for reading and writing data in R.


## Learning Objectives
- getting and setting working directory
- save and load R objects to/from disk
- read and write tabular data
- read data from a url


Factors
-------

A feature is **categorical** if it measures a qualitative category. For
example, the genres `rock`, `blues`, `alternative`, `folk`, `pop` are
categories.

R uses the class `factor` to represent categorical data. Visualizations and
statistical models sometimes treat factors differently than other data types,
so it's important to make sure you have the right data type. If you're ever
unsure, remember that you can check the class of an object with the `class`
function.

When you load a data set, R usually can't tell which features are categorical.
That means identifying and converting the categorical features is up to you.
For beginners, it can be difficult to understand whether a feature is
categorical or not. The key is to think about whether you want to use the
feature to divide the data into groups.

For example, if we want to know how many songs are in the `rock` genre, we
first need to divide the songs by genre, and then count the number of songs in
each group (or at least the `rock` group).

As a second example, months recorded as numbers can be categorical or not,
depending on how you want to use them. You might want to treat them as
categorical (for example, to compute max rainfall in each month) or you might
want to treat them as numbers (for example, to compute the number of months
time between two events).

The bottom line is that you have to think about what you'll be doing in the
analysis. In some cases, you might treat a feature as categorical only for part
of the analysis.

Let's think about which features are categorical in earnings data set. To
refresh our memory of what's in the data set, we can look at the structural
summary:

```{r}
#str(earn)
```

The columns `n_persons` and `median_weekly_earn` are quantitative rather than
categorical, since they measure quantities of people and dollars, respectively.

The `sex`, `race`, `ethnic_origin`, and `age` columns are all categorical,
since they are all qualitative measurements. We can see this better if we use
the `table` function to compute frequencies for the values in the columns:

```{r}
#table(earn$sex)
#
#table(earn$race)
#
#table(earn$ethnic_origin)
#
#table(earn$age)
```

Each column has only a few unique values, repeated many times. These are ideal
for grouping the data. If age had been recorded as a number, rather than a
range, it would probably be better to treat it as quantitative, since there
would be far more unique values. Columns with many unique values don't make
good categorical features, because each group will only have a few elements!

That leaves us with the `year` and `quarter` columns. It's easy to imagine
grouping the data by year or quarter, but these are also clearly numbers. These
columns can be treated as quantitative or categorical data, depending on how we
want to use them to analyze the data.

Let's convert the `age` column to a factor. To do this, use the `factor`
function:
```{r}
#age = factor(earn$age)
#head(age)
```

Notice that factors are printed differently than strings.

The categories of a factor are called _levels_. You can list the levels with
the `levels` function:
```{r}
#levels(age)
```

Factors remember all possible levels even if you take a subset:
```{r}
#age[1:3]
```

This is another way factors are different from strings. Factors "remember" all
possible levels even if they aren't present. This ensures that if you plot a
factor, the missing levels will still be represented on the plot.

You can make a factor forget levels that aren't present with the `droplevels`
function:
```{r}
#droplevels(age[1:3])
```


Special Values
--------------

R has four _special_ values to represent missing or invalid data.

### Missing Values

The value `NA`, called the **missing value**, represents missing entries in a
data set. It's implied that the entries are missing due to how the data was
collected, although there are exceptions. As an example, imagine the data came
from a survey, and respondents chose not to answer some questions. In the data
set, their answers for those questions can be recorded as `NA`.

The missing value is a chameleon: it can be a logical, integer, numeric,
complex, or character value. By default, the missing value is logical, and the
other types occur through coercion (\@ref(implicit-coercion)):

```{r}
class(NA)

class(c(1, NA))

class(c("hi", NA, NA))
```

The missing value is also contagious: it represents an unknown quantity, so
using it as an argument to a function usually produces another missing value.
The idea is that if the inputs to a computation are unknown, generally so is
the output:

```{r}
NA - 3

mean(c(1, 2, NA))
```

As a consequence, testing whether an object is equal to the missing value with
`==` doesn't return a meaningful result:

```{r}
5 == NA

NA == NA
```

You can use the `is.na` function instead:

```{r}
is.na(5)

is.na(NA)

is.na(c(1, NA, 3))
```

Missing values are a feature that sets R apart from most other programming
languages.


### Infinity

The value `Inf` represents infinity, and can be numeric or complex. You're most
likely to encounter it as the result of certain computations:

```{r}
13 / 0

class(Inf)
```

You can use the `is.infinite` function to test whether a value is infinite:

```{r}
is.infinite(3)

is.infinite(c(-Inf, 0, Inf))
```

### Not a Number

The value `NaN`, called **not a number**, represents a quantity that's
undefined mathematically. For instance, dividing 0 by 0 is undefined:

```{r}
0 / 0

class(NaN)
```

Like `Inf`, `NaN` can be numeric or complex.

You can use the `is.nan` function to test whether a value is `NaN`:

```{r}
is.nan(c(10.1, log(-1), 3))
```


### Null

The value `NULL` represents a quantity that's undefined in R. Most of the time,
`NULL` indicates the absence of a result. For instance, vectors don't have
dimensions, so the `dim` function returns `NULL` for vectors:

```{r}
dim(c(1, 2))

class(NULL)

typeof(NULL)
```

Unlike the other special values, `NULL` has its own unique type and class.

You can use the `is.null` function to test whether a value is `NULL`:

```{r}
is.null("null")

is.null(NULL)
```


Indexing
--------

The way to get and set elements of a data structure is by **indexing**.
Sometimes this is also called **subsetting** or (element) **extraction**.
Indexing is a fundamental operation in R, key to reasoning about how to solve
problems with the language.

We first saw indexing in Section \@ref(data-frames), where we used `$`, the
dollar sign operator, to get and set data frame columns. We saw indexing again
in Section \@ref(indexing-vectors), where we used `[`, the indexing or square
bracket operator, to get and set elements of vectors.

The indexing operator `[` is R's primary operator for indexing. It works in
four different ways, depending on the type of the index you use. These four
ways to select elements are:

1. All elements, with no index
2. By position, with a numeric index
3. By name, with a character index
4. By condition, with a logical index

Let's examine each in more detail. We'll use this vector as an example, to keep
things concise:

```{r}
x = c(a = 10, b = 20, c = 30, d = 40, e = 50)
x
```

Even though we're using a vector here, the indexing operator works with almost
all data structures, including factors, lists, matrices, and data frames. We'll
look at unique behavior for some of these later on.

### All Elements

The first way to use `[` to select elements is to leave the index blank. This
selects all elements:
```{r}
x[]
```

This way of indexing is rarely used for getting elements, since it's the same
as entering the variable name without the indexing operator. Instead, its main
use is for setting elements. Suppose we want to set all the elements of `x` to
`5`. You might try writing this:

```{r}
x = 5
x
```

Rather than setting each element to `5`, this sets `x` to the scalar `5`, which
is not what we want. Let's reset the vector and try again, this time using the
indexing operator:

```{r}
x = c(a = 10, b = 20, c = 30, d = 40, e = 50)
x[] = 5
x
```

As you can see, now all the elements are `5`. So the indexing operator is
necessary to specify that we want to set the elements rather than the whole
variable.

Let's reset `x` one more time, so that we can use it again in the next example:

```{r}
x = c(a = 10, b = 20, c = 30, d = 40, e = 50)
```

### By Position

The second way to use `[` is to select elements by position. This happens when
you use an integer or numeric index. We already saw the basics of this in
Section \@ref(indexing-vectors).

The positions of the elements in a vector (or other data structure) correspond
to numbers starting from 1 for the first element. This way of indexing is
frequently used together with the sequence operator `:` to get ranges of
values. For instance, let's get the 2nd through 4th elements of `x`:

```{r}
x[2:4]
```

You can also use this way of indexing to set specific elements or ranges of
elements. For example, let's set the 3rd and 5th elements of `x` to `9` and
`7`, respectively:

```{r}
x[c(3, 5)] = c(9, 7)
x
```

When getting elements, you can repeat numbers in the index to get the same
element more than once. You can also use the order of the numbers to control
the order of the elements:

```{r}
x[c(2, 1, 2, 2)]
```

Finally, if the index contains only negative numbers, the elements at those
positions are excluded rather than selected. For instance, let's get all
elements except the 1st and 5th:

```{r}
x[-c(1, 5)]
```

When you index by position, the index should always be all positive or all
negative. Using a mix of positive and negative numbers causes R to emit error
rather than returning elements, since it's unclear what the result should be:

```{r, error = TRUE}
x[c(-1, 2)]
```

### By Name

The third way to use `[` is to select elements by name. This happens when you
use a character vector as the index, and only works with named data structures.

Like indexing by position, you can use indexing by name to get or set elements.
You can also use it to repeat elements or change the order. Let's get elements
`a`, `c`, `d`, and `a` again from the vector `x`:

```{r}
y = x[c("a", "c", "d", "a")]
y
```

Element names are generally unique, but if they're not, indexing by name gets
or sets the first element whose name matches the index:

```{r}
y["a"]
```

Let's reset `x` again to prepare for learning about the final way to index:

```{r}
x = c(a = 10, b = 20, c = 30, d = 40, e = 50)
```

### By Condition

The fourth and final way to use `[` is to select elements based on a condition.
This happens when you use a logical vector as the index. The logical vector
should have the same length as what you're indexing, and will be recycled if it
doesn't.

#### Congruent Vectors {-}

To understand indexing by condition, we first need to learn about congruent
vectors. Two vectors are **congruent** if they have the same length and they
correspond element-by-element.

For example, suppose you do a survey that records each respondent's favorite
animal and age. These are two different vectors of information, but each person
will have a response for both. So you'll have two vectors that are the same
length:

```{r}
animal = c("dog", "cat", "iguana")
age = c(31, 24, 72)
```

The 1st element of each vector corresponds to the 1st person, the 2nd to the
2nd person, and so on. These vectors are congruent.

Notice that columns in a data frame are always congruent!

#### Back to Indexing {-}

When you index by condition, the index should generally be congruent to the
object you're indexing. Elements where the index is `TRUE` are kept and
elements where the index is `FALSE` are dropped.

If you create the index from a condition on the object, it's automatically
congruent. For instance, let's make a condition based on the vector `x`:

```{r}
is_small = x < 25
is_small
```

The 1st element in the logical vector `is_small` corresponds to the 1st element
of `x`, the 2nd to the 2nd, and so on. The vectors `x` and `is_small` are
congruent.

It makes sense to use `is_small` as an index for `x`, and it gives us all the
elements less than `25`:

```{r}
x[is_small]
```

Of course, you can also avoid using an intermediate variable for the condition:

```{r}
x[x > 10]
```

If you create index some other way (not using the object), make sure that it's
still congruent to the object. Otherwise, the subset returned from indexing
might not be meaningful.

You can also use indexing by condition to set elements, just as the other ways
of indexing can be used to set elements. For instance, let's set all the
elements of `x` that are greater than `10` to the missing value `NA`:

```{r}
x[x > 10] = NA
x
```


### Logic

All of the conditions we've seen so far have been written in terms of a single
test. If you want to use more sophisticated conditions, R provides operators to
negate and combine logical vectors. These operators are useful for working with
logical vectors even outside the context of indexing.

#### Negation {-}

The _NOT operator_ `!` converts `TRUE` to `FALSE` and `FALSE` to `TRUE`:
```{r}
x = c(TRUE, FALSE, TRUE, TRUE, NA)
x
!x
```

You can use `!` with a condition:
```{r}
y = c("hi", "hello")
!(y == "hi")
```

The NOT operator is vectorized.


#### Combinations {-}

R also has operators for combining logical values.

The _AND operator_ `&` returns `TRUE` only when both arguments are `TRUE`. Here
are some examples:

```{r}
FALSE & FALSE
TRUE & FALSE
FALSE & TRUE
TRUE & TRUE
c(TRUE, FALSE, TRUE) & c(TRUE, TRUE, FALSE)
```

The _OR operator_ `|` returns `TRUE` when at least one argument is `TRUE`.
Let's see some examples:

```{r}
FALSE | FALSE
TRUE | FALSE
FALSE | TRUE
TRUE | TRUE
c(TRUE, FALSE) | c(TRUE, TRUE)
```

Be careful: everyday English is less precise than logic. You might say:

> I want all subjects with age over 50 and all subjects that like cats.

But in logic this means:

`(subject age over 50) OR (subject likes cats)`

So think carefully about whether you need both conditions to be true (AND) or
at least one (OR).

Rarely, you might want _exactly one_ condition to be true. The _XOR (eXclusive
OR) function_ `xor()` returns `TRUE` when exactly one argument is `TRUE`. For
example:

```{r}
xor(FALSE, FALSE)
xor(TRUE, FALSE)
xor(TRUE, TRUE)
```

The AND, OR, and XOR operators are vectorized.


#### Short-circuiting {-}

The second argument is irrelevant in some conditions:

* `FALSE &` is always `FALSE`
* `TRUE |` is always `TRUE`

Now imagine you have `FALSE & long_computation()`. You can save time by
skipping `long_computation()`. A _short-circuit operator_ does exactly that.

R has two short-circuit operators:

* `&&` is a short-circuited `&`
* `||` is a short-circuited `|`

These operators only evaluate the second argument if it is necessary to
determine the result. Here are some of these:

```{r}
TRUE && FALSE
TRUE && TRUE
TRUE || TRUE
c(TRUE, FALSE) && c(TRUE, TRUE)
```

For the final expression, notice R only combines the first element of each
vector. The others are ignored. In other words, the short-circuit operators are
_not_ vectorized! Because of this, generally you **should not use** the
short-circuit operators for indexing. Their main use is in writing conditions
for if-statements, which we'll learn about later on.

### Indexing Lists

Lists are a _container_ for other types of R objects. When you select an
element from a list, you can either keep the container (the list) or discard
it. The indexing operator `[` almost always keeps containers.

As an example, let's get some elements from a small list:

```{r}
x = list(first = c(1, 2, 3), second = sin, third = c("hi", "hello"))
y = x[c(1, 3)]
y
class(y)
```

The result is still a list. Even if we get just one element, the result of
indexing a list with `[` is a list:

```{r}
class(x[1])
```

Sometimes this will be exactly what we want. But what if we want to get the
first element of `x` so that we can use it in a vectorized function? Or in a
function that only accepts numeric arguments? We need to somehow get the
element and discard the container.

The solution to this problem is the _extraction operator_ `[[`, which is also
called the _double square bracket operator_. The extraction operator is the
primary way to get and set elements of lists and other containers.

Unlike the indexing operator `[`, the extraction operator always discards the
container:

```{r}
x[[1]]
class(x[[1]])
```

The tradeoff is that the extraction operator can only get or set one element at
a time. Note that the element can be a vector, as above. Because it can only
get or set one element at a time, the extraction operator can only index by
position or name. Blank and logical indexes are not allowed.

The final difference between the index operator `[` and the extraction operator
`[[` has to do with how they handle invalid indexes. The index operator `[`
returns `NA` for invalid vector elements, and `NULL` for invalid list elements:

```{r}
c(1, 2)[10]
x[10]
```

On the other hand, the extraction operator `[[` raises an error for invalid
elements:

```{r, error = TRUE}
x[[10]]
```

The indexing operator `[` and the extraction operator `[[` both work with any
data structure that has elements. However, you'll generally use the indexing
operator `[` to index vectors, and the extraction operator `[[` to index
containers (such as lists).


### Indexing Data Frames

For two-dimensional objects, like matrices and data frames, you can pass the
indexing operator `[` or the extraction operator `[[` a separate index for each
dimension. The rows come first:

```
DATA[ROWS, COLUMNS]
```

For instance, let's get the first 3 rows and all columns of the earnings data:

```{r}
#earn[1:3, ]
```

As we saw in Section \@ref(all-elements), leaving an index blank means all
elements.

As another example, let's get the 3rd and 5th row, and the 2nd and 4th column:

```{r}
#earn[c(3, 5), c(2, 4)]
```

Mixing several different ways of indexing is allowed. So for example, we can
get the same above, but use column names instead of positions:

```{r}
#earn[c(3, 5), c("race", "age")]
```

For data frames, it's especially common to index the rows by condition and the
columns by name. For instance, let's get the `sex`, `age`, and `n_persons`
columns for all rows that pertain to women:

```{r}
#result = earn[earn$sex == "Women", c("sex", "age", "n_persons")]
#head(result)
```

#### The Drop Parameter {-}

If you use two-dimensional indexing with `[` to select exactly one column, you
get a vector:

```{r}
#result = earn[1:3, 2]
#class(result)
```

The container is dropped, even though the indexing operator `[` usually keeps
containers. This also occurs for matrices. You can control this behavior with
the `drop` parameter:

```{r}
#result = earn[1:3, 2, drop = FALSE]
#class(result)
```

The default is `drop = TRUE`.
