Reshaping Tabular Data
======================

Learning Objectives
-------------------

After this lesson, you should be able to:

* Cleaning:
  + Pivot columns in a data set to make it tidy
  + Separate values in a column that contains multiple values per cell
  + Convert columns to appropriate data types


Introduction
------------

This lesson focuses on how to identify untidy tabular data sets and reshape
them to be tidy, in the sense descriped in Section \@ref(tidy-data).

Let's look at some examples of tidy and untidy data sets. The **tidyr** package
provides examples, and as we'll see later, it also provides functions to make
untidy data sets tidy. As usual, we first need to load the package:

```{r}
# install.packages("tidyr")
library(tidyr)
```

Let's start with an example of tidy data. This data set is included in the
**tidyr** package and records the number of tuberculosis cases across several
different countries and years:

```{r}
table1
```

When you first look at a data set, think about what the observations are and
what the features are. If the data set comes with documentation, it may help
you figure this out. Since this data set is a tidy data set, we already know
each row is an observation and each column is a feature.

Features in a data set tend to take one of two roles. Some features are
_identifiers_ that describe the observed subject. These are usually not what
the researcher collecting the data is trying to find out. For example, in the
tuberculosis data set, the `country` and `year` columns are identifiers.

Other features are _measurements_. These are usually the reason the researcher
collected the data. For the tuberculosis data set, the `cases` and `population`
columns are measurements.

Thinking about whether features are identifiers or measurements can be helpful
when you need to use **tidyr** to rearrange a data set.


### Columns into Rows

Tidy data rule 1 says each observation must have its own row. Here's a table
that breaks rule 1:

```{r}
table4a
```

All of the numbers measure the same thing: cases. To make the data tidy, we
must rotate the `1999` and `2000` column names into rows, one for each value in
the columns. The new columns are `year` and `cases`.

This process means less columns (generally) and more rows, so the data set
becomes longer.

We can use the `pivot_longer` function to rotate columns into rows. We need to
specify:

* Columns to rotate as `cols`.
* Name(s) of new identifier column(s) as `names_to`.
* Name(s) of new measuerment column(s) as `values_to`.

Here's the code:
```{r}
pivot_longer(table4a, -country, names_to = "year", values_to = "cases")
```


#### How to Pivot Longer without **tidyr**

You also can do this without **tidyr**:

1. Subset columns to separate `1999` and `2000` into two data frames.
2. Add a `year` column to each.
3. Rename the `1999` and `2000` columns to `cases`.
4. Stack the two data frames with `rbind`.

```{r}
# Step 1
df99 = table4a[-3]
df00 = table4a[-2]

# Step 2
df99$year = "1999"
df00$year = "2000"

# Step 3
names(df99)[2] = "cases"
names(df00)[2] = "cases"

# Step 4
rbind(df99, df00)
```


### Rows into Columns

Tidy data rule 2 says each feature must have its own column. Let's look at a
table that breaks rule 2:

```{r}
table2
```

Here the `count` column contains two different features: cases and population.
To make the data tidy, we must rotate the `count` values into columns, one for
each `type` value. New columns are `cases` and `population`.

This process means less rows and more columns, so the data set becomes wider.

We can use `pivot_wider` to rotate rows into columns. We need to specify:

* Column names to rotate as `names_from`.
* Measurements to rotate as `values_from`.

Here's the code:

```{r}
pivot_wider(table2, names_from = type, values_from = count)
```

#### How to Pivot Wider without **tidyr**

You can also do this without **tidyr**:

1. Subset rows to separate `cases` and `population` values.
2. Remove the `type` column from each.
3. Rename the `count` column to `cases` and `population`.
4. Merge the two subsets by matching `country` and `year`.

```{r}
# Step 1
cases = table2[table2$type == "cases", ]
pop = table2[table2$type == "population", ]

# Step 2
cases = cases[-3]
pop = pop[-3]

# Step 3
names(cases)[3] = "cases"
names(pop)[3] = "population"

# Step 4
tidy = cbind(cases, pop[3])
```

This code uses the `cbind` function to merge the two subsets, but it would be
better to use the `merge` function. The `cbind` function does not use
identifier columns to check that the rows in each subset are from the same
observations.

Run `vignette("pivot")` for more examples of how to use **tidyr**.


### Separating Values

Tidy data rule 3 says each value must have its own cell. Here's a table that
breaks rule 3:

```{r}
table3
```

Cells in the `rate` column contain two values: cases and population. These are
two different features, so to make the data set tidy, we need to separate them
into two different columns.

So how can we separate the `rate` column? The rate column is a character vector
(you can check this with `str(table3)`), so we can use the string processing
functions in the **stringr** package. In particular, we can use the
`str_split_fixed` function:

```{r}
library(stringr)

columns = str_split_fixed(table3$rate, fixed("/"), 2)
```

Now we have a character matrix where the values are in separate columns. Now we
need to combine these with the original data frame. There are several ways to
approach this, but to be safe, let's make a new data frame rather than
overwrite the original. First we make a copy of the original:

```{r}
tidy_tb = table3
```

Next, we need to assign each column in the character matrix to a column in the
`tidy_tb` data frame. Since the columns contain numbers, we can also use the
`as.numeric` function to convert them to the correct data type:

```{r}
tidy_tb$cases = as.numeric(columns[, 1])
tidy_tb$population = as.numeric(columns[, 2])
```

Extracting values, converting to appropriate data types, and then combining
everything into a single data frame is an extremely common pattern in data
science.

Using **stringr** functions is the most general way to separate out values in a
column, but the **tidyr** package also provides a function `separate`
specifically for the case we just worked through. Either package is appropriate
for solving this problem.
