```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Control Flow

:::{.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Create code that only runs when a condition is satisfied
* Identify when a problem requires iteration
* Select appropriate iteration strategies for problems
* Explain what a map (or apply) function does
* Call a function repeatedly with the purrr package's map functions
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following packages:

* dplyr
* ggplot2
* purrr
* readxl
* stringr

@sec-packages explains how to install and load packages.
:::

Suppose you want to read and clean 10 different mass measurement datasets. All
of the datasets have the same structure, so the code to read and clean them is
the same. Based on what you learned in @sec-writing-functions, you decide to
write a function, `read_mass_data`, and reuse the code. But you realize there
are still two problems:

1. When you collected the 8th dataset, you forgot to tare the scale. As a
   result, all of the measurements are 50 grams too high. This is easy to
   correct by subtracting 50, but only the 8th dataset needs this correction.
   You need a way to choose which code to run based on a condition.
2. You still have to write 10 calls to `read_mass_data`. Using a programming
   language is supposed to help you avoid tedious things like this, and this
   wouldn't scale well if you needed to read 100 or 1,000 datasets. You need a
   way to run code multiple times.

Both of these problems are related to the order in which code runs (the flow of
control). R provides several **control flow** expressions that change how code
runs. You can use control flow expressions to choose which code to run based on
a condition or to run code multiple times. Many of the concepts in this chapter
generalize to other programming languages, which provide similar commands to
alter control flow.


## Conditional Expressions

A **conditional** expression is one where the computer must make a decision
about which code to run next. The computer makes its decision by checking
whether a condition is true. In a flowchart, a conditional looks like a branch
(two or more ways to get somewhere):

:::{#fig-conditional}
```{mermaid}
flowchart LR
  start(Start) --> condition{Is condition true?}
    condition -- True --> do_something[Do something]
    condition -- False --> do_something_else["Do something else"]
  do_something --> done(End)
  do_something_else --> done
```

A flowchart that shows a conditional. The computer checks a condition to
decide which code to run.
:::

In most programming languages, the keyword `if` (or some variation of this)
creates a conditional. Because of this, conditionals are also called **if
statements**. In R, the syntax for a conditional is:

```{r}
#| eval: false
if (condition) {
  # This code runs if the condition is TRUE.
} else {
  # This code runs if the condition is FALSE.
}
```

The condition must return a single `TRUE` or `FALSE` value. The `else` part of
the conditional is optional, so if you only want to do something in the true
case, you can just write:

```{r}
#| eval: false
if (condition) {
  # This code runs if the condition is TRUE.
}
```

You can test multiple conditions by replacing the curly braces `{ }` after
`else` with another conditional:

```{r}
#| eval: false
if (condition1) {
  # This code runs if condition1 is TRUE.
} else if (condition2) {
  # This code runs if condition2 is TRUE (and condition1 is FALSE).
} else {
  # This code runs if both condition1 and condition2 are FALSE.
}
```

As an example, suppose we want our code to produce a different greeting
depending on the hour:

* If it's the morning (hours 6-11), the code should produce `"Good morning!"`.
* If it's the afternoon (hours 12-17), the code should produce `"Good
  afternoon!"`.
* For any other hour, the code should produce `"Hello."`.

We can use a conditional to do this (see @sec-logic if you need a refresher on
how to write conditions that return `TRUE` and `FALSE` values):

```{r}
hour = 10

if (hour >= 6 && hour <= 11) {
  greeting = "Good morning!"
} else if (hour >= 12 && hour <= 17) {
  greeting = "Good afternoon!"
} else {
  greeting = "Hello."
}

greeting
```

Notice how changing the value of `hour` changes the result:

```{r}
hour = 13

if (hour >= 6 && hour <= 11) {
  greeting = "Good morning!"
} else if (hour >= 12 && hour <= 17) {
  greeting = "Good afternoon!"
} else {
  greeting = "Hello."
}

greeting
```

Try out a few values for `hour` yourself. Then try modifying the code to give
another greeting, `"Good evening!"` if it's the evening (hours 18-19).

:::{.callout-note}
In R, conditionals automatically return the value on the last line that runs
(similar to functions). So another way to write the greeter example above is:

```{r}
hour = 13

greeting = if (hour >= 6 && hour <= 11) {
  "Good morning!"
} else if (hour >= 12 && hour <= 17) {
  "Good afternoon!"
} else {
  "Hello."
}

greeting
```
:::

:::{.callout-note}
You can **nest** one conditional inside the curly braces of another. Nesting is
useful when you want to check a condition, do some computations, and then check
another condition under the assumption that the first condition was `TRUE`.
:::

:::{.callout-warning}
Conditionals correspond to **special cases** in your code. Having lots of
special cases can make code harder to understand and maintain.

It's not bad to use conditionals, but be mindful of how you use them. Think
about whether there's a more general way to do things instead. Be especially
wary of nested conditionals.
:::


## Case Study: U.S. Alternative Fueling Stations, Part II {#sec-case-study-fuel-2}

Let's continue the U.S. Alternative Fueling Stations case study from
@sec-case-study-fuel-1. In the previous part, we wrote a function,
`read_fuel_sheet`, to read and clean sheets from the dataset Excel file:

```{r}
#| output: false
library("dplyr")
library("readxl")
library("stringr")

read_fuel_sheet = function(path, year) {
  sheet = as.character(year)
  stations = read_excel(path, sheet = sheet, skip = 1)

  # Clean up the column names.
  names = names(stations)
  is_dot_name = str_starts(names, fixed("..."))
  names[is_dot_name] = as.character(stations[2, is_dot_name])
  names = str_to_lower(names)
  names(stations) = names

  # Remove blank rows.
  stations = filter(stations, !is.na(biodiesel))

  # Correct column types and add year column.
  stations$electric = as.numeric(stations$electric)
  # TODO: hydrogen, propane
  stations$year = year

  stations
}
```

We designed the function around the 2007 data sheet, but the format of the
sheets changed in 2014. So the `read_fuel_sheet` function emits an error for
2014 and later. Let's edit the function so that it works well for every year.
We can use conditionals for code that should only run for some of the years.

To get started, let's write code to read and clean the data for 2023, since
it's the last year. We can use the code in `read_fuel_sheet` as a reference.
We'll still use the readxl package's `read_excel` function to read the sheet:

```{r}
path = "data/2007-2023_us_alt_fuels.xlsx"
stations = read_excel(path, sheet = "2023", skip = 1)
head(stations)
```

This immediately reveals one reason why the `read_fuel_sheet` function didn't
work. For 2014 and later, all of the column names are in the 2nd row of the
sheet. There's no need to get names from other rows. Let's make a note of this
in a comment.

We still need to convert the column names to lowercase:

```{r}
names = names(stations)
# For 2014 and later, no need to get column names from two different rows.
names = str_to_lower(names)
```

One of the columns is named `renewable diesel`, which will be difficult to use
in analysis because it contains a space. We can use stringr's `str_replace_all`
to replace the space with an underscore:

```{r}
names = str_replace_all(names, fixed(" "), "_")
names
```

Four of the names seem incorrect: `electrica`, `hydrogenb`, `propanec`, and
`totald`. If we examine the dataset with spreadsheet software, we can see that
these column names have alphabetic superscripts in 2014 and later. There are
many ways to fix this, but we'll use dplyr's `rename` function:

```{r}
names(stations) = names

# For 2014 and later:
stations = rename(
  stations,
  electric = electrica,
  hydrogen = hydrogenb,
  propane = propanec,
  total = totald
)
head(stations)
```

The next step in `read_fuel_sheet` is to remove blank rows (rows with missing
values). The 2023 data frame has rows that are partially blank, but they do
have values in the `electric` column. By inspecting the sheet, we can see that
for 2014 and later, the `electric` column contains multiple values separated by
the pipe character `|`. The `electric` values for each state are also split
across two rows. Since we want to compare the data over time, we only need the
station count, which is in the first row. So we'll use the code from
`read_fuel_sheet` that filters out the blank rows:

```{r}
stations = filter(stations, !is.na(biodiesel))
head(stations)
```

With that done, let's split the `electric` column into two columns, since it
contains two values. The first value is the number of stations, which we'll
keep in the `electric` column. The second value is the number of outlets, which
we'll put in a column called `electric_outlets`. To make it easier to convert
the columns to numbers, we'll remove all of the commas (`,`) with stringr's
`str_replace_all` function. Then we can use stringr's `str_split_fixed`
function to split the column in two. Finally, we can use `as.numeric` to
convert the columns to numbers before saving them back into the data frame:

```{r}
# For 2014 and later:
electric = str_replace_all(stations$electric, fixed(","), "")
electric = str_split_fixed(electric, fixed(" | "), 2)
stations$electric = as.numeric(electric[, 1])
stations$electric_outlets = as.numeric(electric[, 2])
```

The `hydrogen` and `propane` columns have the same problem. We'll skip fixing
them, since the code to do so is almost identical.

The 2023 data frame is now clean enough for analysis. The code to read and
clean it is:

```{r}
#| eval: false
stations = read_excel(path, sheet = "2023", skip = 1)

names = names(stations)
# For 2014 and later, no need to get column names from two different rows.
names = str_to_lower(names)
names = str_replace_all(names, fixed(" "), "_")
names(stations) = names

# For 2014 and later:
stations = rename(
  stations,
  electric = electrica,
  hydrogen = hydrogenb,
  propane = propanec,
  total = totald
)

stations = filter(stations, !is.na(biodiesel))

# For 2014 and later:
electric = str_replace_all(stations$electric, fixed(","), "")
electric = str_split_fixed(electric, fixed(" | "), 2)
stations$electric = as.numeric(electric[, 1])
stations$electric_outlets = as.numeric(electric[, 2])
```

We can use conditionals to combine this with code in the body of the
`read_fuel_sheet` function. Specifically, we need to check `year < 2014` or
`year >= 2014` at a few points. One way to write the code for the combined
function is:

```{r}
read_fuel_sheet = function(path, year) {
  sheet = as.character(year)
  stations = read_excel(path, sheet = sheet, skip = 1)

  # Clean up the column names.
  names = names(stations)
  if (year < 2014) {
    is_dot_name = str_starts(names, fixed("..."))
    names[is_dot_name] = as.character(stations[2, is_dot_name])
  }
  names = str_to_lower(names)
  names = str_replace_all(names, fixed(" "), "_")
  names(stations) = names
  if (year >= 2014) {
    stations = rename(
      stations,
      electric = electrica,
      hydrogen = hydrogenb,
      propane = propanec,
      total = totald
    )
  }

  # Remove blank rows.
  stations = filter(stations, !is.na(biodiesel))

  # Correct column types and add year column.
  if (year >= 2014) {
    electric = str_replace_all(stations$electric, fixed(","), "")
    electric = str_split_fixed(electric, fixed(" | "), 2)
    stations$electric = as.numeric(electric[, 1])
    stations$electric_outlets = as.numeric(electric[, 2])
    # TODO: hydrogen, propane
  }
  stations$electric = as.numeric(stations$electric)
  # TODO: hydrogen, propane
  stations$year = year

  stations
}
```

Notice how using multiple conditionals makes it harder to understand what the
function does. We could instead use one conditional, but we'd have to repeat
parts of the code that are the same. Alternatively, we could move some of the
cleaning steps to other functions, so that `read_fuel_sheet` is shorter and
easier to understand (without changing what it does).

Let's test the new `read_fuel_sheet` function on the 2023 data:

```{r}
read_fuel_sheet(path, 2023)
```

We should also make sure that it still returns the same result for the 2007
data:

```{r}
read_fuel_sheet(path, 2007)
```

This looks good. We've now got a function that can read any of the years. In
part III of this case study (@sec-case-study-fuel-3), we'll see how to use the
function to read all of them.


## Iteration

R is powerful tool for automating tasks that have repetitive steps. For
example, you can:

* Apply a transformation to an entire column of data.
* Compute distances between all pairs from a set of points.
* Read a large collection of files from disk in order to combine and analyze
  the data they contain.
* Simulate how a system evolves over time from a specific set of starting
  parameters.
* Scrape data from the pages of a website.

You can implement concise, efficient solutions for these kinds of tasks by
using **iteration**, which means repeating a computation many times. In a
flowchart, iteration looks like a loop (a path that begins and ends at the same
place):

:::{#fig-while-loop}
```{mermaid}
flowchart LR
  start(Start) --> condition{Is condition true?}
    condition -- True --> body[Do something]
    body --> condition
    condition -- False ----> done(End)
```

A flowchart that shows a while loop (one kind of iteration). The computer
checks a condition at the beginning of each iteration to decide whether to
continue iterating or stop.
:::

R provides four different strategies for writing iterative code:

1. Vectorization, where a function is implicitly called on each element of a
   vector. This was introduced in @sec-vectorization.
2. Map (or apply) functions, where a function is explicitly called on each
   element of a data structure.
3. Loops, where an expression is evaluated repeatedly until some condition is
   met.
4. Recursion, where a function calls itself.

Vectorization is the most efficient and concise iteration strategy, but also
the least flexible, because it only works with specific functions and with
vectors. Map functions are more flexible---they work with any function and any
data structure with elements---but less efficient and less concise. Loops and
recursion provide the most flexibility but are the least concise. Recursion
tends to be the least efficient iteration strategy in R.

The next two sections, @sec-the-purrr-package and @sec-for-loops, introduce map
functions and loops, respectively. @sec-how-to-write-iterative-code explains
how to choose which iteration strategy to use and how to write iterative code.


## The purrr Package {#sec-the-purrr-package}

@sec-vectorization introduced vectorization, a convenient and efficient way to
compute multiple results. That section also mentioned that some of R's
functions---the ones that summarize or aggregate data---are not vectorized.

The `class` function is an example of a function that's not vectorized. If we
call the `class` function on the least terns dataset, we get just one result
for the dataset as a whole:

```{r}
class(terns)
```

We can get the class of a single column by selecting the column with `$`, the
dollar sign operator:

```{r}
class(terns$year)
```

What if we want the classes of all the columns? We could write a call to
`class` for each column, but that would be tedious. When you're working with a
programming language, you should try to avoid tedium; there's usually a better,
more automated way.

Data frames are technically lists (@sec-lists), where each column is one
element. With that in mind, what we need here is a line of code that calls
`class` on each element of the data frame. The idea is similar to
vectorization, but since we have a list and a non-vectorized function, we have
to do a bit more than just call `class(terns)`.

The [purrr][] package is a collection of functions to help you do things
repeatedly or for each element of a data structure. Install and load the
package in order to follow along:

[purrr]: https://purrr.tidyverse.org/

```{r}
# install.packages("purrr")
library("purrr")
```

The package's `map` function calls a function on each element of a vector
or list. We sometimes also say it **maps** or **applies** a function over the
elements. The syntax is:

```{r}
#| eval: false
map(DATA, F, ...)
```

The `map` function calls the function `F` once for each element of `DATA`. It
passes the element to `F` as the first argument. It also passes the `...`
arguments to `F`, which are constant across all of the calls.

Let's try this out with the least terns data and the `class` function:

```{r}
map(terns, class)
```

The result is similar to if the `class` function was vectorized. In fact, if we
use a vector and a vectorized function with `map`, the result is nearly
identical to the result from vectorization:

```{r}
x = c(1, 2, pi)

sin(x)

map(x, sin)
```

The only difference is that the result from `map` is a list. In fact, the `map`
function always returns a list with one element for each element of the input
data.


### Other Map Functions

The purrr package provides many different map functions, all of which have
names that start with `map`. All of them call another function on each element
of a data structure. They also all have the same syntax. Where they differ is
in how they return results. A few of these are shown in @tbl-map-functions.

Function  | Return Type
--------- | -----------
`map_lgl` | logical
`map_int` | integer
`map_dbl` | numeric (double)
`map_chr` | character
`map`     | list

: {#tbl-map-functions}


:::{.callout-note title="Note: Apply Functions"}
R's **apply functions** are a built-in equivalent to map functions. The
`lapply`, `sapply`, and `tapply` functions are the three most important
functions in the family of apply functions, but there are many more. The
`lapply` function is nearly identical to the `map` function.

We focus on and recommend the map functions rather than the apply functions
because they are more consistent in their syntax and specific in their return
types. You can learn more about R's apply functions by reading [this
StackOverflow post][apply].

[apply]: https://stackoverflow.com/a/7141669
:::

<!--
When you have a choice between using vectorization or a map function, you
should always choose vectorization. Vectorization is clearer---compare the two
lines of code above---and it's also significantly more efficient. In fact,
vectorization is the most efficient way to call a function repeatedly in R.

As we saw with the `class` function, there are some situations where
vectorization is not possible. That's when you should think about using a map
function.
-->

Let's look at some examples of the other map functions. If we use `map_chr` to
find the classes of the columns in the least terns data, we get a character
vector:

```{r}
map_chr(terns, class)
```

Likewise, if we use `map_dbl` to compute the `sin` values, we get a numeric
vector, the same as from vectorization:

```{r}
map_dbl(x, sin)
```

In spite of that, vectorization is still more efficient than `sapply`, so use
vectorization instead when possible.

<!--
Map functions are incredibly useful for summarizing data. For example, suppose
we want to compute the frequencies for all of the columns in the least terns
dataset that aren't numeric.

First, we need to identify the columns. One way to do this is with the
`is.numeric` function. Despite the name, this function actually tests whether
its argument is a real number, not whether it its argument is a numeric vector.
In other words, it also returns true for integer values. We can use `map_lgl`
to apply this function to all of the columns in the least terns dataset:

```{r}
is_not_number = !map_lgl(terns, is.numeric)
is_not_number
```

Is it worth using R code to identify the non-numeric columns? Since there are
only `r ncol(terns)` columns in the least terns dataset, maybe not. But if the
dataset was larger, with say 100 columns, it definitely would be.

In general, it's a good habit to use R to do things rather than do them
manually. You'll get more practice programming, and your code will be more
flexible if you want to adapt it to other datasets.

Now that we know which columns are non-numeric, we can use the `table` function
to compute frequencies. We only want to compute frequencies for those columns,
so we need to subset the data:

```{r}
map(terns[, is_not_number], table)
```

We use `map` rather than one of the other map functions for this step because
the table for each column will have a different length.
-->

The [purrr documentation][purrr] provides more details about how to use the
many functions in the package.



## For Loops {#sec-for-loops}

A **for loop** runs the code in its body once for each element of a data
structure. In a flowchart, a for loop looks like a loop (a path that begins and
ends in the same place)---hence the name:

:::{#fig-for-loop}
```{mermaid}
flowchart LR
  start(Start) --> condition{Are there more elements?}
    condition -- Yes --> get_elt["Get the next element"]
    get_elt --> body[Do something]
    body --> condition
    condition -- No ----> done(End)
```

A flowchart that shows a map function, apply function, or for loop. At the
beginning of each iteration, the computer checks whether there is another
element in the data to decide whether to continue iterating. If there is, the
computer automatically assigns the next element to a variable.
:::

In most programming languages, the keyword `for` creates a for loop. In R, the
syntax of a for loop is:

```{r}
#| eval: false
for (element in data) {
  # This code runs once for each element in data.
}
```

The loop automatically assigns the next element of `data` to the variable
`element` at the beginning of each iteration. The loop iterates once for each
element, unless a keyword causes the loop to end early.

:::{.callout-tip}
For loops and map functions do almost the same thing. Here's how the for loop
syntax above translates into a map function:

```{r}
#| eval: false
map(data, \(element) {
  # This code runs once for each element in data.
})
```

Map functions tend to be easier to use because they automatically return the
result from each iteration. For loops don't, and leave it up to you to figure
out how to store the results.

The limitation of map functions is that the each iteration must be independent.
You can't have an iteration that depends on the result of a prior iteration.
For loops don't have this limitation. Dependent iterations are common in
simulations, but not so common in data cleaning and analysis tasks.
:::

As a demonstration, let's print out a message with the current iteration number
for 5 iterations:

```{r}
for (i in 1:5) {
  message("Hi from iteration ", i)
}
```

Unlike other iteration strategies, loops don't return a result automatically.
It's up to you to use variables to store any results you want to use later. To
do this:

1. Before the loop, create a result vector (or list) that's the same length as
   the number of iterations. Fill vector with 0s or some other placeholder
   value. You can use functions such as `integer`, `numeric`, and `character`
   to do this.
2. In the loop, use indexing to replace the elements of the result vector as
   they're computed.

This approach to storing results is called **pre-allocation**. Here's an
example of pre-allocation for a loop that computes what happens if you
repeatedly call the sine function on a value:

```{r}
# Number of iterations:
n = 1 + 99

# Create a numeric vector with n elements and set the first element to 1.
result = numeric(n)
result[1] = 1

for (i in 2:n) {
  # Get the result from the previous iteration.
  # This is why this loop starts at i = 2.
  prev_result = result[i - 1]

  # Compute the sine of the previous result.
  # Save it into the results vector at position i.
  result[i] = sin(prev_result)
}
```

Plotting the result vector makes it easier to see the pattern in the values:

```{r}
library("ggplot2")

result_df = data.frame(x = 1:n, y = result)
ggplot(result_df) + geom_point() + aes(x = x, y = y)
```

:::{.callout-note}
The `break` keyword causes a loop to immediately exit. It only makes sense to
use `break` inside of a conditional (otherwise, the loop will exit in the first
iteration).

The `next` keyword causes a loop to immediately go to the next iteration. As
with `break`, it only makes sense to use `next` inside of a conditional.
:::

:::{.callout-note title="Note: While Loops"}
A **while loop** makes the computer check a condition at the beginning of each
iteration to decide whether to run the code in the loop's body. @fig-while-loop
shows what a while loop looks like in a flowchart.

In most programming languages, the keyword `while` creates a while loop. In R,
the syntax of a while loop is:

```{r}
#| eval: false
while (condition) {
  # This code runs repeatedly until the condition is FALSE.
}
```

While loops are a generalization of for loops, and only do the bare minimum
necessary to iterate. They tend to be most useful when you don't know how many
iterations are necessary to complete a task.
:::


## How to Write Iterative Code {#sec-how-to-write-iterative-code}

At first it might seem difficult to decide if and what kind of iteration to
use. Start by thinking about whether you need to do something over and over. If
you don't, then you probably don't need to use iteration. If you do, then try
iteration strategies in this order:

1. Vectorization
2. Map (or apply) functions
    * Try an apply function if iterations are independent.
3. Loops
    * Try a for-loop if some iterations depend on others.
    <!--
    * Try a while-loop if the number of iterations is unknown.
    -->
4. Recursion (which isn't covered here)
    * Convenient for naturally recursive tasks (like Fibonacci), but often
      there are faster solutions.

Start by writing the code for just one iteration. Make sure that code works;
it's easy to test code for one iteration.

When you have one iteration working, then try using the code with an iteration
strategy (you will have to make some small changes). If it doesn't work, try to
figure out which iteration is causing the problem. One way to do this is to use
`message` to print out information. Then try to write the code for the broken
iteration, get that iteration working, and repeat this whole process.


## Case Study: U.S. Alternative Fueling Stations, Part III {#sec-case-study-fuel-3}

Let's finish the U.S. Alternative Fueling Stations case study we began in
@sec-case-study-fuel-1 and continued in @sec-case-study-fuel-2. In the previous
part, we modified the `read_fuel_sheet` function so that it can read and clean
sheets for any year in the dataset Excel file:

```{r}
read_fuel_sheet = function(path, year) {
  sheet = as.character(year)
  stations = read_excel(path, sheet = sheet, skip = 1)

  # Clean up the column names.
  names = names(stations)
  if (year < 2014) {
    is_dot_name = str_starts(names, fixed("..."))
    names[is_dot_name] = as.character(stations[2, is_dot_name])
  }
  names = str_to_lower(names)
  names = str_replace_all(names, fixed(" "), "_")
  names(stations) = names
  if (year >= 2014) {
    stations = rename(
      stations,
      electric = electrica,
      hydrogen = hydrogenb,
      propane = propanec,
      total = totald
    )
  }

  # Remove blank rows.
  stations = filter(stations, !is.na(biodiesel))

  # Correct column types and add year column.
  if (year >= 2014) {
    electric = str_replace_all(stations$electric, fixed(","), "")
    electric = str_split_fixed(electric, fixed(" | "), 2)
    stations$electric = as.numeric(electric[, 1])
    stations$electric_outlets = as.numeric(electric[, 2])
    # TODO: hydrogen, propane
  }
  stations$electric = as.numeric(stations$electric)
  # TODO: hydrogen, propane
  stations$year = year

  stations
}
```

In this part, let's use the function to read all of the sheets, then combine
them into a single data frame and make a plot of number of electric stations
over time. Imagine what it would look like if we wrote out the code to read the
data for each year separately:

```{r}
#| eval: false
path = "data/2007-2023_us_alt_fuels.xlsx"
read_fuel_sheet(path, 2007)
read_fuel_sheet(path, 2008)
# ...
read_fuel_sheet(path, 2023)
```

The year is what differs between the calls, so we need to iterate over the
years. We could create a vector of years with `2007:2023`. Instead, let's use
readxl's `excel_sheets` function to get the sheet names, which correspond to
years, and convert them to numbers:

```{r}
years = excel_sheets(path)
# Skip the first sheet (which is documentation) and convert to numbers.
years = as.numeric(years[-1])
years
```

Now we need to call the `read_fuel_sheet` function with each year in the
`years` vector. We can use purrr's `map` function to do this. Since the year
should be the second argument to `read_fuel_sheet`, we have to write the call
out explicitly (alternatively, we could modify `read_fuel_sheet` so that the
year is the first argument):

```{r}
#| output: false
sheets = map(years, \(year) {
  read_fuel_sheet(path, year)
})
```

This gives a list of data frames, one for each year. We want to bind them into
a single data frame, stacking the rows of each one atop the next. We can use
dplyr's `bind_rows` function to do this:

```{r}
stations = bind_rows(sheets)
head(stations)
```

With the data in good shape, we're ready to make a plot. Let's make a line plot
with the year on the x-axis, the number of electric stations on the y-axis, and
the lines color-coded by state:

```{r}
ggplot(stations) +
  geom_line() +
  aes(x = year, y = electric, color = state)
```

This plot is a bit hard to read because there are so many states. Let's use
dplyr's `filter` function to focus only on states on the west coast:

```{r}
west = filter(
  stations,
  state %in% c("California", "Oregon", "Washington", "Alaska", "Hawaii")
)

ggplot(west) +
  geom_line() +
  aes(x = year, y = electric, color = state)
```

Now we can see that California has substantially more electric stations than
the other west coast states. This could be due to its larger population or due
to increased funding for alternative fueling stations.


## Case Study: CA Hospital Utilization

The California Department of Health Care Access and Information (HCAI) requires
hospitals in the state to submit detailed information each year about how many
beds they have and the total number of days for which each bed was occupied.
The HCAI publishes the data to the [California Open Data Portal][data.ca.gov].
Let's use R to read data from 2016 to 2023 and investigate whether hospital
utilization is noticeably different in and after 2020.

[data.ca.gov]: https://data.ca.gov/

The dataset consists of a separate Microsoft Excel file for each year. Before
2018, HCAI used a data format (in Excel) called ALIRTS. In 2018, they started
collecting more data and switched to a data format called SIERA. The 2018 data
file contains a **crosswalk** that shows the correspondence between SIERA
columns and ALIRTS columns.

:::{.callout-important}
[Click here][ca-hospitals] to download the CA Hospital Utilization dataset (8
Excel files).

If you haven’t already, we recommend you create a directory for this workshop.
In your workshop directory, create a `data/ca_hospitals` subdirectory. Download
and save the dataset in the `data/ca_hospitals` subdirectory.

[ca-hospitals]: https://ucdavis.box.com/s/g5tanw22647dw3uyhv3om0zt3duuj5lc
:::

When you need to solve a programming problem, get started by writing some
comments that describe the problem, the inputs, and the expected output. Try to
be concrete. This will help you clarify what you're trying to achieve and serve
as a guiding light while you work.

As a programmer (or any kind of problem-solver), you should always be on the
lookout for ways to break problems into smaller, simpler steps. Think about
this when you frame a problem. Small steps are easier to reason about,
implement, and test. When you complete one, you also get a nice sense of
progress towards your goal.

For the CA Hospital Utilization dataset, our goal is to investigate whether
there was a change in hospital utilization in 2020. Before we can do any
investigation, we need to read the files into R. The files all contain tabular
data and have similar formats, so let's try to combine them into a single data
frame. We'll say this in the framing comments:

```{r}
# Read the CA Hospital Utilization dataset into R. The inputs are yearly Excel
# files (2016-2023) that need to be combined. The pre-2018 files have a
# different format from the others. The result should be a single data frame
# with information about bed and patient counts.
#
# After reading the dataset, we'll investigate utilization in 2020.
```

"Investigate utilization" is a little vague, but for an exploratory data
analysis, it's hard to say exactly what to do until you've started working with
the data.

We need to read multiple files, but we can simplify the problem by starting
with just one. Let's start with the 2023 data. It's in an Excel file, which you
can read with the `read_excel` function from the readxl package. If it's your
first time using the readxl package, you'll need to install it:

```{r}
#| eval: false
install.packages("readxl")
```

The `read_excel` function requires the path to the file as the first argument.
You can optionally provide the sheet name or number (starting from 1) as the
second argument. Open up the Excel file in your computer's spreadsheet program
and take a look. There are multiple sheets, and the data about beds and
patients are in the second sheet. Back in R, read just the second sheet:

```{r}
library("readxl")

path = "data/ca_hospitals/hosp23_util_data_final.xlsx"
sheet = read_excel(path, sheet = 2)
head(sheet)
```

The first four rows contain metadata about the columns. The first hospital,
Alameda Hospital, is listed in the fifth row. So let's remove the first four
rows:

```{r}
sheet = sheet[-(1:4), ]
head(sheet)
```

Some datasets also have metadata in the last rows, so let's check for that
here:

```{r}
tail(sheet)
```

Sure enough, the last row contains what appears to be a count of the hospitals
rather than a hospital. Let's remove it by calling `head` with a negative
number of elements, which removes that many elements from the end:

```{r}
sheet = head(sheet, -1)
tail(sheet)
```

There are a lot of columns in sheet, so let's make a list of just a few that
we'll use for analysis. We'll keep:

* Columns with facility name, location, and operating status
* All of the columns whose names start with `TOT`, because these are totals for
  number of beds, number of census-days, and so on.
* Columns about acute respiratory beds, with names that contain `RESPIRATORY`,
  because they might also be relevant.

We can use the stringr package, which provides string processing functions, to
help us get the `TOT` and `RESPIRATORY` column names. If it's your first time
using the stringr package, you'll have to install it:

```r
install.packages("stringr")
```

We can use stringr's `str_starts` function to check whether column names start
with `TOT` and its `str_detect` function to check whether column names contain
`RESPIRATORY`:

```{r}
library("stringr")

facility_cols = c(
  "FAC_NAME", "FAC_CITY", "FAC_ZIP", "FAC_OPERATED_THIS_YR", "FACILITY_LEVEL",
  "TEACH_HOSP", "COUNTY", "PRIN_SERVICE_TYPE"
)

cols = names(sheet)
tot_cols = cols[str_starts(cols, "TOT")]
respiratory_cols = cols[str_detect(cols, "RESPIRATORY")]
```

The `TOT` and `RESPIRATORY` columns all contain numbers, but the element type
is `character`, so let's cast to numbers them with `as.numeric`. We'll also add
a column with the year:

```{r}
numeric_cols = c(tot_cols, respiratory_cols)
sheet[numeric_cols] = lapply(sheet[numeric_cols], as.numeric)

sheet$year = 2023
```

Now we'll select only the columns we identified as useful:

```{r}
sheet = sheet[c("year", facility_cols, numeric_cols)]

head(sheet)
```

Lowercase names are easier to type, so let's also make all of the names
lowercase with stringr's `str_to_lower` function:

```{r}
names(sheet) = str_to_lower(names(sheet))
head(sheet)
```

We've successfully read one of the files! Since the 2018-2023 files all have
the same format, it's likely that we can use almost the same code for all of
them. Any time you want to reuse code, it's a sign that you should write a
function, so that's what we'll do. We'll take all of the code we have so far
and put it in the body of a function called `read_hospital_data`, adding some
comments to indicate the steps:

```r
read_hospital_data = function() {
  # Read the 2nd sheet of the file.
  path = "data/ca_hospitals/hosp23_util_data_final.xlsx"
  sheet = read_excel(path, sheet = 2)

  # Remove the first 4 and last row.
  sheet = sheet[-(1:4), ]
  sheet = head(sheet, -1)

  # Select only a few columns of interest.
  facility_cols = c(
    "FAC_NAME", "FAC_CITY", "FAC_ZIP", "FAC_OPERATED_THIS_YR",
    "FACILITY_LEVEL", "TEACH_HOSP", "COUNTY", "PRIN_SERVICE_TYPE"
  )

  cols = names(sheet)
  tot_cols = cols[str_starts(cols, "TOT")]
  respiratory_cols = cols[str_detect(cols, "RESPIRATORY")]

  numeric_cols = c(tot_cols, respiratory_cols)
  sheet[numeric_cols] = lapply(sheet[numeric_cols], as.numeric)

  sheet$year = 2023

  sheet = sheet[c("year", facility_cols, numeric_cols)]

  # Rename the columns to lowercase.
  names(sheet) = str_to_lower(names(sheet))

  sheet
}
```

As it is, the function still only reads the 2023 file. The other files have
different paths, so the first thing we need to do is make the `path` variable a
parameter. We'll also make a `year` parameter, for the year value inserted as a
column:

```{r}
read_hospital_data = function(path, year) {
  # Read the 2nd sheet of the file.
  sheet = read_excel(path, sheet = 2)

  # Remove the first 4 and last row.
  sheet = sheet[-(1:4), ]
  sheet = head(sheet, -1)

  # Select only a few columns of interest.
  facility_cols = c(
    "FAC_NAME", "FAC_CITY", "FAC_ZIP", "FAC_OPERATED_THIS_YR",
    "FACILITY_LEVEL", "TEACH_HOSP", "COUNTY", "PRIN_SERVICE_TYPE"
  )

  cols = names(sheet)
  tot_cols = cols[str_starts(cols, "TOT")]
  respiratory_cols = cols[str_detect(cols, "RESPIRATORY")]

  numeric_cols = c(tot_cols, respiratory_cols)
  sheet[numeric_cols] = lapply(sheet[numeric_cols], as.numeric)

  sheet$year = year

  sheet = sheet[c("year", facility_cols, numeric_cols)]

  # Rename the columns to lowercase.
  names(sheet) = str_to_lower(names(sheet))

  sheet
}
```

Test the function out on a few of the files to make sure it works correctly:

```{r}
head(
  read_hospital_data(
    "data/ca_hospitals/hosp23_util_data_final.xlsx", 2023
  )
)
```

```{r}
head(
  read_hospital_data(
    "data/ca_hospitals/hosp21_util_data_final-revised-06.15.2023.xlsx", 2021
  )
)
```

The function appears to work correctly for two of the files, so let's work
towards trying it on all of the 2018-2023 files. We can use the built-in
`list.files` function to get the paths to the files by setting `full.names =
TRUE` (otherwise it just returns the names of the files):

```{r}
paths = list.files("data/ca_hospitals/", full.names = TRUE)
paths
```

In addition to the file paths, we also need the year for each file.
Fortunately, the last two digits of the year are included in each file's name.
We can write a function to get these. We'll use R's built-in `basename`
function to get the file names from the paths, and stringr's `str_sub` function
to get a substring (the year) from the name:

```{r}
get_hospital_year = function(path) {
  name = basename(path)
  year = str_sub(name, 5, 6)
  as.integer(year) + 2000
}

get_hospital_year(paths[1])
```

<!--
Now we can use a for-loop to iterate over all of the files. We'll skip the
pre-2018 files for now:

```{r}
hosps = list()

for (i in seq_along(paths)) {
  path = paths[[i]]
  year = get_hospital_year(path)
  # Only years 2018-2023.
  if (year >= 2018) {
    hosp = read_hospital_data(path, year)
    hosps[[i]] = hosp
  }
}

length(hosps)
```
-->

With that done, we need to read the pre-2018 files. In the 2018 file, the
fourth sheet is a crosswalk that shows which columns in the pre-2018 files
correspond to columns in the later files. Let's write some code to read the
crosswalk. First, read the sheet:

```{r}
path = "data/ca_hospitals/hosp18_util_data_final.xlsx"
cwalk = read_excel(path, sheet = 4)
head(cwalk)
```

The new and old column names are in the fourth and fifth columns, respectively,
so we'll get just those:

```{r}
cwalk = cwalk[, 4:5]
head(cwalk)
```

Finally, let's turn the `cwalk` data frame into a named vector. We'll make the
old column names the names and the new column names the elements. This way we
can easily look up the new name for any of the old columns by indexing. Some of
the column names in `cwalk` have extra spaces at the end, so we'll use
stringr's `str_trim` function to remove them:

```{r}
cwalk_names = str_trim(cwalk[[2]])
cwalk = str_trim(cwalk[[1]])
names(cwalk) = cwalk_names

head(cwalk)
```

We can now define a new version of the `read_hospital_data` function that uses
the crosswalk to change the column names when `year < 2018`. Let's also change
function to exclude columns with `ALOS` in the name, because they have no
equivalent in the pre-2018 files:


```{r}
read_hospital_data = function(path, year) {
  # Read the 2nd sheet of the file.
  sheet = read_excel(path, sheet = 2)

  # Remove the first 4 and last row.
  sheet = sheet[-(1:4), ]
  sheet = head(sheet, -1)

  # Fix pre-2018 column names.
  if (year < 2018) {
    new_names = cwalk[names(sheet)]
    new_names[is.na(new_names)] = names(sheet)[is.na(new_names)]
    names(sheet) = new_names
  }

  # Select only a few columns of interest.
  facility_cols = c(
    "FAC_NAME", "FAC_CITY", "FAC_ZIP", "FAC_OPERATED_THIS_YR",
    "FACILITY_LEVEL", "TEACH_HOSP", "COUNTY", "PRIN_SERVICE_TYPE"
  )

  cols = names(sheet)
  tot_cols = cols[str_starts(cols, "TOT")]
  respiratory_cols = cols[str_detect(cols, "RESPIRATORY")]

  numeric_cols = c(tot_cols, respiratory_cols)
  numeric_cols = numeric_cols[!str_detect(numeric_cols, "ALOS")]
  sheet[numeric_cols] = lapply(sheet[numeric_cols], as.numeric)

  sheet$year = year

  sheet = sheet[c("year", facility_cols, numeric_cols)]

  # Rename the columns to lowercase.
  names(sheet) = str_to_lower(names(sheet))

  sheet
}
```

Now we can test the function on all of the files. We'll use a for-loop to
iterate over all of the paths, read the data for each one, and store the result
in a list:

```{r}
hosps = list()

for (i in seq_along(paths)) {
  path = paths[[i]]
  year = get_hospital_year(path)
  hosp = read_hospital_data(path, year)
  hosps[[i]] = hosp
}

length(hosps)
```

We can use the `do.call` and `rbind` functions to bind the rows, or stack, the
list of data frames:

```{r}
hosps = do.call(rbind, hosps)
head(hosps)
```

We've finally got all of the data in a single data frame!

To begin to address whether hospital utilization changed in 2020, let's make a
bar plot of total census-days:

```{r}
library("ggplot2")

(
    ggplot(hosps) +
    aes(x = year, weight = tot_cen_days) +
    geom_bar()
)
```

According to the plot, total census-days was slightly lower in 2020 than in
2019. This is a bit surprising, but it's possible that California hospitals
typically operate close to maximum capacity and were not able to substantially
increase the number of beds in 2020 in response to the COVID-19 pandemic. You
can use other columns in the dataset, such as `tot_lic_beds` or
`tot_lic_bed_days`, to check this.

Let's also look at census-days for acute respiratory care:


```{r}
(
    ggplot(hosps) +
    aes(x = year, weight = acute_respiratory_care_cen_days) +
    geom_bar()
)
```

In this plot, there's a clear uptick in census-days in 2020, and then an
interesting decrease to below 2019 levels in the years following. Again, you
could use other columns in the dataset to investigate this further. We'll end
this case study here, having accomplished the difficult task of reading the
data and the much easier task of doing a cursory preliminary analysis of the
data.


## Case Study: The Collatz Conjecture

The [Collatz Conjecture][collatz] is a conjecture in math that was introduced
in 1937 by Lothar Collatz and remains unproven today, despite being relatively
easy to explain. Here's a statement of the conjecture:

[collatz]: https://en.wikipedia.org/wiki/Collatz_conjecture

> Start from any positive integer. If the integer is even, divide by 2. If the
> integer is odd, multiply by 3 and add 1.
>
> If the result is not 1, repeat using the result as the new starting value.
>
> The result will always reach 1 eventually, regardless of the starting value.

The sequences of numbers this process generates are called **Collatz
sequences**. For instance, the Collatz sequence starting from 2 is `2, 1`. The
Collatz sequence starting from 12 is `12, 6, 3, 10, 5, 16, 8, 4, 2, 1`.

You can use iteration to compute the Collatz sequence for a given starting
value. Since each number in the sequence depends on the previous one, and since
the length of the sequence varies, a while-loop is the most appropriate
iteration strategy:

```{r}
n = 5
i = 0
while (n != 1) {
  i = i + 1
  if (n %% 2 == 0) {
    n = n / 2
  } else {
    n = 3 * n + 1
  }
  message(n, " ", appendLF = FALSE)
}
```

As of 2020, scientists have used computers to check the Collatz sequences for
every number up to approximately $2^{64}$.


::: {.callout-note title="See also"}
For more details about the Collatz Conjecture, check out [this
video][collatz-video].

[collatz-video]: https://www.youtube.com/watch?v=094y1Z2wpJg
:::
